### 2023-09-28 doc/_variational_inference_primer.qmd
fix high-dimensional integral

In this section, we will briefly introduce the idea behind variational
inference. Recall that any Bayesian inference problem deals with the joint
distribution between observations $\underline{x}$ and unobserved latent
variables $\underline{\theta}$. This joint distribution can be written as the
product of a distribution of the observations $\underline{x}$ conditioned on the
$\underline{\theta}$ and the marginal distribution of these latent variables,
i.e.,
$$
\pi(\underline{x}, \underline{\theta}) =
\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
$${#eq-vi_joint_dist}
A Bayesian inference pipeline's objective is to compute the latent variables'
posterior probability given a set of observations. This computation is
equivalent to updating our prior beliefs about the set of values that the latent
variables take after taking in new data. We write this as Bayes theorem
$$
\pi(\underline{\theta} \mid \underline{x}) = 
\frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }.
$${#eq-vi_bayes_theorem}
The main technical challenge for working with @eq-vi_bayes_theorem comes from
the computation of the denominator, also known as the *evidence* or the
*marginalized likelihood*. The reason computing this term is challenging is
because it involves a (potentially) high-dimensional integral of the form
$$
\pi(\underline{x}) = 
\int\cdots\int d\underline{\theta}^K \pi(\underline{x}, \underline{\theta}) = 
\int\cdots\int d\underline{\theta}^K \pi(\underline{x} \mid \underline{\theta})
\pi(\underline{\theta}),
$${#eq-vi_evidence_int}
where $K$ is the dimesionality of the $\underline{\theta}$ vector. Here, the
integrals are taken over the support---the set of values valid for the
distribution---of $\pi(\underline{\theta})$. However, only a few selected
distributions have a closed analytical form; thus, in most cases
@eq-vi_evidence_int must be solved numerically.

@@

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d\underline{\theta}^K
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d\underline{\theta}^K$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d\underline{\theta}^K q_\phi(\underline{\theta}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

1. The proposed approximate distribution must be normalized, i.e.,
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta}) = 1.
$${#eq-vi_q_norm}
2. The law of the unconscious statistician (LOTUS) establishes that for any
probability density function, it must be true that
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta})
f(\underline{\theta}) = \left\langle 
    f(\underline{\theta}) 
\right\rangle_{q_\phi},
$${#eq-vi_lotus}
where $\left\langle\cdot\right\rangle_{q_\phi}$ is the expected value over the
$q_\phi$ distribution.

>>>>

In this section, we will briefly introduce the idea behind variational
inference. Recall that any Bayesian inference problem deals with the joint
distribution between observations $\underline{x}$ and unobserved latent
variables $\underline{\theta}$. This joint distribution can be written as the
product of a distribution of the observations $\underline{x}$ conditioned on the
$\underline{\theta}$ and the marginal distribution of these latent variables,
i.e.,
$$
\pi(\underline{x}, \underline{\theta}) =
\pi(\underline{x} \mid \underline{\theta}) \pi(\underline{\theta}).
$${#eq-vi_joint_dist}
A Bayesian inference pipeline's objective is to compute the latent variables'
posterior probability given a set of observations. This computation is
equivalent to updating our prior beliefs about the set of values that the latent
variables take after taking in new data. We write this as Bayes theorem
$$
\pi(\underline{\theta} \mid \underline{x}) = 
\frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }.
$${#eq-vi_bayes_theorem}
The main technical challenge for working with @eq-vi_bayes_theorem comes from
the computation of the denominator, also known as the *evidence* or the
*marginalized likelihood*. The reason computing this term is challenging is
because it involves a (potentially) high-dimensional integral of the form
$$
\pi(\underline{x}) = 
\int\cdots\int d^K\underline{\theta}\; \pi(\underline{x}, \underline{\theta}) = 
\int\cdots\int d^K\underline{\theta}\; 
\pi(\underline{x} \mid \underline{\theta})
\pi(\underline{\theta}),
$${#eq-vi_evidence_int}
where $K$ is the dimesionality of the $\underline{\theta}$ vector. Here, the
integrals are taken over the support---the set of values valid for the
distribution---of $\pi(\underline{\theta})$. However, only a few selected
distributions have a closed analytical form; thus, in most cases
@eq-vi_evidence_int must be solved numerically.

@@

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d^K\underline{\theta}\;
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d^K\underline{\theta}\;$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d^K\underline{\theta}\; q_\phi(\underline{\theta}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

1. The proposed approximate distribution must be normalized, i.e.,
$$
\int d^K\underline{\theta}\; q_\phi(\underline{\theta}) = 1.
$${#eq-vi_q_norm}
2. The law of the unconscious statistician (LOTUS) establishes that for any
probability density function, it must be true that
$$
\int d^K\underline{\theta}\; q_\phi(\underline{\theta})
f(\underline{\theta}) = \left\langle 
    f(\underline{\theta}) 
\right\rangle_{q_\phi},
$${#eq-vi_lotus}
where $\left\langle\cdot\right\rangle_{q_\phi}$ is the expected value over the
$q_\phi$ distribution.
